---
title: "Trực quan hóa dữ liệu với R"
author: "Cuong Sai"
output:
  html_document:
    css: style.css
    code_folding: show
    fig_caption: yes
    highlight: textmate
    theme: spacelab
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: true
subtitle: '[Khoa học dữ liệu với R](https://svcuong.github.io/)'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,      # Output code chunks
    message = FALSE,  # Toggle off message output 
    warning = FALSE)  # Toggle off warning output
```


# Giới thiệu về data visualization và ggplot2

Trực quan hóa dữ liệu (`data visualization`) đóng vai trò vô cùng quan trọng trong phân tích dữ liệu và là phương pháp tiếp cận mạnh nhất giúp diễn giải nhanh chóng và hiệu quả dữ liệu thô thành các thông tin có ích hỗ trợ cho quá trình đưa ra quyết định. 

Trong `R` đã tích hợp rất nhiều các hàm `base` cho phép  thiết kế và vẽ một biểu đồ một cách nhanh chóng và đẹp mắt. Bên cạnh những hàm `base`, `R` cũng có những gói  được phát triển chuyên dụng cho data visualization, nổi bật hơn cả trong số đó phải kể đến gói  `ggplot2` trong hệ sinh thái `tidyverse`. 


**ggplot2** là một gói đồ hòa mạnh mẽ được viết bởi [Hadley Wickham]() với mục đích tạo ra các **elegant graphics**. Hai chữ cái đầu `gg` trong `ggplot2` có nghĩa là **Grammar of Graphics** - tức là ngữ pháp biểu đồ.  Nguyên lý ngữ pháp biểu đồ trong `ggplot2` được mô tả như sau:

```
Plot = data + Aesthetics + Geometry
```

**Trong đó:**

- data: dữ liệu ở dạng data frame

- Aesthetics: sử dụng để quy định các biến cho trục Ox và Oy. Ngoài ra có thể sử dụng để điều chỉnh màu, kích cỡ, hình dạng,... của biểu đồ.

- Geometry: Thiết lập kiểu biểu đồ mà cần thể hiện (như histogram, box plot, line plot, density plot, dot plot,..)


**Ngoài ra có thể thêm các lớp không bắt buộc như:**

- Tính toán thống kê (statítical transformation): Sử dụng các tính toán trong biểu đồ. VD: Thêm các biểu đồ xu hướng dạng trơ (geom_smooth) , số lượng bin trong histogram… Lớp này có thể nằm lồng trong các lớp khác)

- Tỷ lệ (scale): Xác định tỷ lệ trong biểu đồ

- Trục tọa độ (Coordinate): Xác định hệ trục tọa độ vẽ biểu đồ. VD: Hệ tọa độ 3 trục trong không gian (hệ tọa đồ Descartes), hệ tọa độ cực

- Hệ thống Facet (tạm dịch: cách thức sắp xếp số liệu trên cùng 1 biểu đồ): Xác định cách thức sắp xếp vị trí của nhiều biểu đồ trên cùng một màn hình.

- Các yêu tố khác: Tên biểu đồ, tên trục, chú giải,


**Hai hàm chính được sử dụng để tạo biểu đồ trong ggplot2 là:

- `qplot()`: Hàm tạo biểu đồ nhanh, dễ sử dụng dành cho các biểu đồ đơn giản

- `qqplot()`: Hàm linh hoạt và mạnh mẽ hơn `qplot()` để xây dựng biểu đồ bằng các lớp chồng lên nhau.

**cài đặt và load ggplot2:**

```{r, eval=FALSE}
# Installation
install.packages('ggplot2')
# Loading
library(ggplot2)
```

Nếu các bạn đã biết sử dụng các biểu đồ base trên R thì ở bài này sẽ từng bước so sánh giữa base graphics với ggplot2, từ đó giúp bạn chuyển dần sang sử dụng ggplot2 một cách hiệu quả nhất.


# Line graph

## Making a Basic Line Graph

```{r}
library(ggplot2)
df <- pressure
ggplot(df, aes(x=temperature, y=pressure)) + geom_line()
```

## Adding Points to a Line Graph

```{r}
ggplot(df, aes(x=temperature, y=pressure)) + geom_line() + geom_point()
```

## Log axis

```{r}
ggplot(df, aes(x=temperature, y=pressure)) + geom_line() + geom_point() + scale_y_log10()
```


## Changing the Appearance of Lines

```{r}
ggplot(df, aes(x=temperature, y=pressure)) + geom_line(linetype="dashed", size=1, colour="blue")
```


## Changing the Appearance of Points

```{r}
ggplot(df, aes(x=temperature, y=pressure)) + geom_line() + geom_point(size=4, shape=22, colour="darkred", fill="pink")
```


## Making a Line Graph with Multiple Lines

```{r}
library(dplyr)
nmonths = 24
x = seq(as.Date("2015/1/1"), by = "month", length.out = nmonths)
df1 <- data.frame(dates = x,Variable = rnorm(mean = 0.75,nmonths))
df2 <- data.frame(dates = x,Variable = rnorm(mean = -0.75,nmonths))
df3 <- data.frame(dates = x,Variable = rnorm(mean = 0.3,nmonths))
df <- df1 %>%  mutate(cat = "A") %>%  bind_rows(df2 %>% mutate(cat = "B")) %>%  bind_rows(df3 %>% mutate(cat = "C"))
head(df)
```

```{r}
# Tạo multiple lines graphs
ggplot(df, aes(x = dates, y = Variable, color = cat)) +  geom_line() 
```
```{r}
ggplot(df, aes(x = dates, y = Variable, linetype = cat)) +  geom_line() 
```


## Making a Graph with a Shaded Area

```{r}
# Convert the sunspot.year data set into a data frame for this example
sunspotyear <- data.frame(
Year = as.numeric(time(sunspot.year)),
Sunspots = as.numeric(sunspot.year)
)
ggplot(sunspotyear, aes(x=Year, y=Sunspots)) + geom_area()
```


```{r}
ggplot(sunspotyear, aes(x=Year, y=Sunspots)) +
geom_area(colour="black", fill="blue", alpha=.2)
```

```{r}
ggplot(sunspotyear, aes(x=Year, y=Sunspots)) +
geom_area(fill="blue", alpha=.2) +
geom_line()
```

## Making a Stacked Area Graph

```{r}
library(gcookbook) # For the data set
ggplot(uspopage, aes(x=Year, y=Thousands, fill=AgeGroup)) + geom_area()
```


```{r}
ggplot(uspopage, aes(x=Year, y=Thousands, fill=AgeGroup)) +
geom_area(colour="black", size=.2, alpha=.4) +
scale_fill_brewer(palette="Blues", breaks=rev(levels(uspopage$AgeGroup)))
```

## Making a Proportional Stacked Area Graph

```{r}
library(gcookbook) # For the data set
library(plyr) # For the ddply() function
# Convert Thousands to Percent
uspopage_prop <- ddply(uspopage, "Year", transform,
Percent = Thousands / sum(Thousands) * 100)
```


```{r}
ggplot(uspopage_prop, aes(x=Year, y=Percent, fill=AgeGroup)) +
geom_area(colour="black", size=.2, alpha=.4) +
scale_fill_brewer(palette="Blues", breaks=rev(levels(uspopage$AgeGroup)))
```


## Adding a Confidence Region

```{r}
library(gcookbook) # For the data set
# Grab a subset of the climate data
clim <- subset(climate, Source == "Berkeley", select=c("Year", "Anomaly10y", "Unc10y"))
```

```{r}
# Shaded region
ggplot(clim, aes(x=Year, y=Anomaly10y)) +
geom_ribbon(aes(ymin=Anomaly10y-Unc10y, ymax=Anomaly10y+Unc10y),
alpha=0.2) +
geom_line()
```


```{r}
# With a dotted line for upper and lower bounds
ggplot(clim, aes(x=Year, y=Anomaly10y)) +
geom_line(aes(y=Anomaly10y-Unc10y), colour="grey50", linetype="dotted") +
geom_line(aes(y=Anomaly10y+Unc10y), colour="grey50", linetype="dotted") +
geom_line()
```


# Bar Graphs

## Making a Basic Bar Graph

```{r}
library(gcookbook) # For the data set
ggplot(pg_mean, aes(x=group, y=weight)) + geom_bar(stat="identity")
```

```{r}
ggplot(BOD, aes(x=Time, y=demand)) + geom_bar(stat="identity")
```

```{r}
# Convert Time to a discrete (categorical) variable with factor()
ggplot(BOD, aes(x=factor(Time), y=demand)) + geom_bar(stat="identity")
```

```{r}
ggplot(pg_mean, aes(x=group, y=weight)) +
geom_bar(stat="identity", fill="lightblue", colour="black")
```
```{r}
pg_mean
```


## Grouping Bars Together

```{r}
library(gcookbook) # For the data set
cabbage_exp
```



# Sử dụng ggplot2 cho dữ liệu chuỗi thời gian

```{r}
library(ggplot2)
# Dữ liệu demo: economics trong gói ggplot2
head(economics)
```

## Tạo line plots  cơ bản:

```{r}
# Basic line plot
ggplot(data = economics, aes(x = date, y = pop))+
  geom_line(color = "#00AFBB", size = 2)
# Plot a subset of the data
ss <- subset(economics, date > as.Date("2006-1-1"))
ggplot(data = ss, aes(x = date, y = pop)) + 
  geom_line(color = "#FC4E07", size = 2)
```

## Điều chỉnh kích thước line

```{r}
ggplot(data = economics, aes(x = date, y = pop)) +
  geom_line(aes(size = unemploy/pop), color = "#FC4E07")
```

## Tạo multiple time series plots

```{r}
# Để tạo multiple plot bởi 2 biến psavert và uempmed theo dates. Đầu tiên cần định hình lại data sửu dụng tidyr package
library(tidyr)
library(dplyr)
df <- economics %>%
  select(date, psavert, uempmed) %>%
  gather(key = "variable", value = "value", -date)
head(df, 3)
```

```{r}
# Multiple line plot
ggplot(df, aes(x = date, y = value)) + 
  geom_line(aes(color = variable), size = 1) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  theme_minimal()
```

```{r}
# Area plot
ggplot(df, aes(x = date, y = value)) + 
  geom_area(aes(color = variable, fill = variable), 
            alpha = 0.5, position = position_dodge(0.8)) +
  scale_color_manual(values = c("#00AFBB", "#E7B800")) +
  scale_fill_manual(values = c("#00AFBB", "#E7B800"))
```

## Set date axis limits

Key R function: scale_x_date()
```{r}
# Base plot with date axis
p <- ggplot(data = economics, aes(x = date, y = psavert)) + 
     geom_line(color = "#00AFBB", size = 1)
p
# Set axis limits c(min, max)
min <- as.Date("2002-1-1")
max <- NA
p + scale_x_date(limits = c(min, max))
```

## Định dạngt date axis labels

Key function: scale_x_date().

To format date axis labels, you can use different combinations of days, weeks, months and years:

- Weekday name: use %a and %A for abbreviated and full weekday name, respectively
- Month name: use %b and %B for abbreviated and full month name, respectively
- %d: day of the month as decimal number
- %Y: Year with century.
- See more options in the documentation of the function ?strptime

```{r}
# Format : month/year
p + scale_x_date(date_labels = "%b/%Y")
```

## Add trend smoothed line

Key function: stat_smooth()

```{r}
p + stat_smooth(
  color = "#FC4E07", fill = "#FC4E07",
  method = "loess"
  )
```

## ggplot2 mở rộng cho đối tượng ts

The ggfortify package is an extension to ggplot2 that makes it easy to plot time series objects (Horikoshi and Tang 2017). It can handle the output of many time series packages, including: zoo::zooreg(), xts::xts(), timeSeries::timSeries(), tseries::irts(), forecast::forecast(), vars:vars().

Another interesting package is the ggpmisc package (Aphalo 2017), which provides two useful methods for time series object:

- stat_peaks() finds at which x positions local y maxima are located, and
- stat_valleys() finds at which x positions local y minima are located.

Here, we’ll show how to easily:

- Visualize a time series object, using the data set AirPassengers (monthly airline passenger numbers 1949-1960).
- Identify shifts in mean and/or variance in a time series using the changepoint package.
- Detect jumps in a data using the strucchange package and the data set Nile (Measurements of the annual flow of the river Nile at Aswan).
- Detect peaks and valleys using the ggpmisc package and the data set lynx (Annual Canadian Lynx trappings 1821–1934).

First, install required R packages:
```{r, eval=FALSE}
install.packages(
  c("ggfortify", "changepoint",
    "strucchange", "ggpmisc")
)
```

Then use the autoplot.ts() function to visualize time series objects, as follow:

```{r}
library(ggfortify)
library(magrittr) # for piping %>%
# Plot ts objects
autoplot(AirPassengers)
# Identify change points in mean and variance
AirPassengers %>%
  changepoint:: cpt.meanvar() %>%  # Identify change points
  autoplot()
# Detect jump in a data
strucchange::breakpoints(Nile ~ 1) %>%
  autoplot()
```

Detect peaks and valleys:

```{r}
library(ggpmisc)
ggplot(lynx, as.numeric = FALSE) + geom_line() + 
  stat_peaks(colour = "red") +
  stat_peaks(geom = "text", colour = "red", 
             vjust = -0.5, x.label.fmt = "%Y") +
  stat_valleys(colour = "blue") +
  stat_valleys(geom = "text", colour = "blue", angle = 45,
               vjust = 1.5, hjust = 1,  x.label.fmt = "%Y")+
  ylim(-500, 7300)
```

**Tài liệu tham khảo:**

http://www.sthda.com/english/articles/32-r-graphics-essentials/128-plot-time-series-data-using-ggplot/

Aphalo, Pedro J. 2017. Ggpmisc: Miscellaneous Extensions to ’Ggplot2’. https://CRAN.R-project.org/package=ggpmisc.

Horikoshi, Masaaki, and Yuan Tang. 2017. Ggfortify: Data Visualization Tools for Statistical Analysis Results. https://CRAN.R-project.org/package=ggfortify.
